\bf Model Visualization Techniques for a Social Network Model

### 2017-09-20 paper.Rnw
so very very very close

<<pkgs>>=
library(tidyverse)
library(RSiena)
library(network)
library(sna)
library(geomnet)
library(GGally)
library(netvizinf)
library(RColorBrewer)
library(gridExtra)
library(cowplot)
library(extrafont)
loadfonts(quiet = T)

@@

\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states. This needs more fleshing out. Adding to the to-do list.}

\subsubsection{Continuous Time Markov Chain (CTMC)}
\st{
In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$. This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero. For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states, $2^{n(n-1)}$: there are two possible states for an edge, and there are $n(n-1)$ edge relationships (excluding self-ties).  The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$. Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state. Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row. Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged. All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.

@@

\subsection{The Models}

@@

In Section~\ref{sec:samemoddiffdat}, we also fit models M1, M2, and M3 to the senate collaboration data for comparison. %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}. 
In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:

\begin{align*}
f_i(x)^{M4} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_5s_{i5} \\
f_i(x, z)^{M5} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_6s_{i6}.
\end{align*}

In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect. This parameter attempts to measure the effect of a node being completely isolated in the network. If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated. In M5,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect. The values of the covariate are converted to numeric, $\{1,2,3\}$, respectively, and mean-centered within \texttt{RSiena}. The mean-centered values of $z \in \{\text{Democrat, Independent, Republican}\}$ are $\{-0.96, 0.04, 1.04\}$. The parameter $\beta_6$ attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value. If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties. \st{Due to the intractability of SAOMs, it is difficult to know for certain how to interpret a fitted value of a parameter. We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible. But,} by exploring these models more visually, we aim to understand these effects, their interpretations, and the model fitting process better.

@@

clintonSenate %>% filter(source %in% target)

@@

<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=3, fig.cap="Significant effects for the two data sets, at the level of 0.10 as calculated by the Wald-type test available in the SIENA software.">>=
smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
smfr_effs$inter1[is.na(smfr_effs$inter1)] <- "none"
library(ggrepel)
set.seed(12345678)
smfr_effs %>% filter(type == "eval" & Waldpval <=0.10) %>% 
ggplot(aes(x = estimate, y = Waldpval, color = inter1)) + 
  #geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +  
  geom_vline(xintercept = 0) + 
  geom_point(size = 3) + 
  geom_text_repel(aes(label = shortName)) + 
  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") + 
  ggtitle("Significant effects at level 0.10 for friendship data") + 
  coord_flip() + 
  ThemeNoNet
sen_effs <- read_csv("data/congress/senateModelSigEffects.csv")
#sen_effs$Waldpval <- parse_number(sen_effs$Waldpval)
sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, color = inter1)) + 
  # geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +  
  geom_vline(xintercept = 0) + 
  geom_point(size = 3) + 
  geom_text_repel(aes(label = shortName)) + 
  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") + 
  ggtitle("Significant effects at level 0.10 for senate data") + 
  coord_flip() + 
  ThemeNoNet
@

\st{The \texttt{RSiena} manual contains over eighty possible effects to include in the model. In order to select parameters to include in the models for our example data, we searched through the possible effects available to model given the data structure to find \textit{significant} effects. We tested for significance using the Wald-type tests built into \texttt{RSiena} for one-at-a-time effects testing. We start with the outdegree and reciprocity measures as the foundation of the models we fit, then add one evaluation effect, fit the model, test the additional effect for significance, and repeat for all possible parameters to add to the model. We performed this procedure for both the small friendship and the senate collaboration data. The results for significant effects are shown in Figure~\ref{fig:sigtesting}. We see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten. In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less. This may suggest that larger data sets tend to result in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test.}

@@

We have varied model settings already by choosing models M1, M2, M3 to fit to our small firendship network data set. In Section~\ref{sec:samemodsamedat}, we fit these models to the data 1,000 times, and in this section, we explore simulations from these three models given the mean values of from the 1,000 fitted parameter values as the parameters in our models. From each of these three models using the means of parameter estimates as our fixed parameter values, we simulated 1,000 observations from each of the three models. In this process, we condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models with the given parameter values. From these simulations, we first create a visualization that represents an average network.

@@

To do create the average network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times. As in Figure~\ref{fig:summnetM1}, edges only appear in the average network if they appear more than 5\% of the time in the simulations. In Figure~\ref{fig:averagenet}, we show the first wave, the ``average'' network from the three models we fit, and the second wave. \st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1. Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3. In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2. None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated. In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7. XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.

@@

\subsubsection{Fitting the same model to different data}\label{sec:samemoddiffdat}
% table of fitted M1 M2 M3
\begin{table}
\scalebox{0.9}{
\begin{tabular}{l|ccc|ccc }
  \hline
  & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
  \hline
  & M1 & M2 & M3 & M1 & M2 & M3 \\
  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
  $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
  \hline
\end{tabular}
}
\caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network. Each model was fit 1,000 times to the friend data, while each model was fit 100 times to the senate data.}
\end{table}
\subsubsection{Fitting the same model to the same data}\label{sec:samemodsamedat}

@@

@@

To further investigate this potentially problematic relationship between the parameter values, we look at correlations between each of the parameter estimates in each model. In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models. The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3.

<<getdata>>=
nulls <- read.csv("data/distribution_null_model.csv")
alt <- read.csv("data/distribution_jumpTT_model.csv")
alt2nd <- read.csv("data/distribution_dblpairs_model.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
null_mod_sv <- as.numeric(nulls[,-c(1,6,7)] %>% summarise_each(funs(mean)))
alt_mod_sv <- as.numeric(alt[,-c(1, 7,8)] %>% summarise_each(funs(mean)))
alt_mod2_sv <- as.numeric(alt2nd[,-c(1,7,8)] %>% summarise_each(funs(mean)))
@

<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs. The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.height=4>>=

#simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")

@@

@@

\st{The last principle of model visualization is to explore the process of fitting the model, instead of just focusing on the end result. This principle is perhaps the most important for SAOMs because} the model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user. Hiding the MCMC steps is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data, obtaining parameter estimates, \st{and drawing conclusions or making predictions}. We are more interested in \textit{how} the models are fit, so we extracted and explored the different steps of that process.

A key component of each step of the SIENA method of moments algorithm is the ``microstep" process. A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{m-1}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$. This microstep process stops when the simulated network has acheived the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where 
$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
This simulation process follows the steps of the continuous-time Markov chain. \st{Each tie change in the CTMC is referred to as one ``microstep". At each microstep}, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Saving and exploring all of these steps is not computationally efficient if one is only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is exactly what we did to create our visualizations. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations. We wanted to view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

The first vizualization we present here is an animation of microsteps simulated to emulate the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure ~\ref{fig:alldataviz}. Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}. When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears. If there are no changes at a particular microstep, no changes are seen. Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie can be viewed at INSERT LINK TO GIF.

@@

set.seed(34569234)
create_net_animate(static_pall) + 
  facet_wrap(~.frame, nrow=2, labeller = 'label_both') + 
  ThemeNet + 
  theme(panel.background = element_rect(color = 'black'))
@

The top row of Figure~\ref{fig:makinggif} shows an edge being removed, and the bottom row shows one being added. In both cases, the ego nodes chosen to act change color from black to red, and they also increase greatly in size. In the case of an edge being removed, shown in the top row, the edge that currently exists is emphasized with the same color and size change that the node gets, and as the animation proceeds the edge shrinks to nothing, as the ego node shrinks and changes color back to its original black. If an edge is being added, as in the bottom row of the figure, the ego node's appearances changes in the sampe way as when the edge is being removed, whil the edge appears colored red from nothing, and grows to a large size, then changes color and size to match the rest of the edges, while the node shrinks and changes color to match the other nodes.

We also use animation to view the changing structure of the adjacency matrix the microsteps. The adjacency matrices for the three waves of friendship data as shown in Figure~\ref{fig:alldataviz} are ordered by node id. There are 16 nodes in the data, numbered 1-16, and that order is used on the $x$ and $y$ axes for the matrix visualization. Viewing the adjacency matrices with this arbitrary ordering does not provide much information to the viewer about the underlying structure of the network. This lack of perceived structure would be exacerbated in an animation, so we adjust the ordering so that the viewer can perceive more of the structure of the graph. This process is known as matrix seriation \citep{seriation}.

>>>>

<<pkgs>>=
library(tidyverse)
library(RSiena)
library(network)
library(sna)
library(geomnet)
library(GGally)
devtools::install_github("sctyner/netvizinf")
library(netvizinf)
library(RColorBrewer)
library(gridExtra)
library(cowplot)
library(extrafont)
loadfonts(quiet = T)

@@

%\hh{XXX This paragraph ends a bit abruptly - what is the future state probability used for?} \st{XX The future state probability is used within the continuous time markov chain to move through the different network states. This needs more fleshing out. Adding to the to-do list.}

\subsubsection{Continuous Time Markov Chain (CTMC)}
\st{
In the continuous-time Markov chain literature, see for instance \citet{ctmc}, chains are characterized by their \textit{generator} or \textit{intensity} matrix $\mathbf{Q}$. This matrix describes the rate of change between two states of the CTMC process, and the rows of this matrix always add to zero. For directed networks with binary edge variables like the ones we will be working with, there are a very large number of possible states for a directed network with $n$ nodes. We denote the state space as $\mathcal{X}$, a set which contains $2^{n(n-1)}$ states: there are two possible states for an edge, $\{0,1\}$, and there are $n(n-1)$ edge relationships because the network is directed and we exclude self-ties.  The intensity matrix for a CTMC in a SAOM is then a square matrix of dimension $2^{n(n-1)} \times 2^{n(n-1)}$. Only one tie changes at a time in the CTMC, resulting in $n(n-1)$ reachable states from the current network state. Thus, the intensity matrix $\mathbf{Q}$ is very sparse, with only $n(n-1) + 1$ non-zero entries in each row. Note that $n(n-1)$ of these entries represent the possible states that are one edge different from a given state, while the additional non-zero entry is for the state to remain unchanged. All other entries in a row are structural zeroes because those network states cannot be reached from the current state in a single change.

@@

\subsection{The Models}\label{sec:models}

@@

In Section~\ref{sec:samemoddiffdat}, we also fit models M1, M2, and M3 to the senate collaboration data for comparison. \st{For fitting M2 to the senate data, we use the number of bills authored by each senator as the node covariate for the jumping transitive triplet variable. In terms of the senate data, then, the value of $\beta_3$ should dictate how willing a senator is to coauthor a bill on which the author of the bill has a different level of authorship, assuming there is an intermediary between the two who has the same authorship level as the first.} %Comparison of the parameter estimates of these models fit to the two different data sets is given in Table~\ref{tab:meanests}. 
%In addition, we fit two more models, M4 and M5 to the senate data that were found to be significant according to the Wald-type test:

%\begin{align*}
%f_i(x)^{M4} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_5s_{i5} \\
%f_i(x, z)^{M5} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_6s_{i6}.
%\end{align*}

%In M4, $s_{i5} = \mathbb{I}(x_{i+} = x_{+i} = 0)$ is an indicator variable for isolated nodes, called the network-isolate effect. This parameter attempts to measure the effect of a node being completely isolated in the network. If the value of the parameter $\beta_5$ is high, nodes are encouraged to remain isolated. In M5,  $s_{i6} = \sum\limits_j x_{ij}x_{ji}\mathbb{I}(z_i = z_j)$ is a measure of reciprocity between senators in the same party, $z_i, z_j \in \{\text{Democrat, Independent, Republican}\}$, generally called the same covariate reciprocity effect. The values of the covariate are converted to numeric, $\{1,2,3\}$, respectively, and mean-centered within \texttt{RSiena}. The mean-centered values of $z \in \{\text{Democrat, Independent, Republican}\}$ are $\{-0.96, 0.04, 1.04\}$. The parameter $\beta_6$ attempts to measure the propensity for forming reciprocal ties between nodes with the same covariate value. If the value of the parameter $\beta_6$ is high, nodes belonging to the same covariate group are encourage to reciprocate incoming ties. 
\st{Due to the intractability of SAOMs, it is difficult to know for certain how to interpret a fitted value of a parameter. We can make educated guesses based on the definition of the effect and the sign of the fitted value, but a direct interpretation is not always possible. But,} by exploring these models more visually, we aim to understand these effects, their interpretations, and the model fitting process better.

@@

@@

<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.">>=
smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
smfr_effs$inter1[is.na(smfr_effs$inter1)] <- "none"
library(ggrepel)
set.seed(12345678)
smfr_effs %>% filter(type == "eval" & Waldpval <=0.10) %>% 
ggplot(aes(x = estimate, y = Waldpval, color = inter1)) + 
  #geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +  
  geom_vline(xintercept = 0) + 
  geom_point(size = 3) + 
  #geom_text_repel(aes(label = shortName)) + 
  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") + 
  ggtitle("Significant effects at level 0.10 for friendship data") + 
  coord_flip() + 
  ThemeNoNet
sen_effs <- read_csv("data/congress/senateModelSigEffects.csv")
#sen_effs$Waldpval <- parse_number(sen_effs$Waldpval)
sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, color = inter1)) + 
  # geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +  
  geom_vline(xintercept = 0) + 
  geom_point(size = 3) + 
  #geom_text_repel(aes(label = shortName)) + 
  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") + 
  ggtitle("Significant effects at level 0.10 for senate data") + 
  coord_flip() + 
  ThemeNoNet
@

\st{The \texttt{RSiena} manual contains over eighty possible effects to include in the model. In order to select parameters to include in the models for our example data, we searched through the possible effects available to model given the data structure to find \textit{significant} effects. We tested for significance using the Wald-type tests built into \texttt{RSiena} for one-at-a-time effects testing. We start with the outdegree and reciprocity measures as the foundation of the models we fit, then add one evaluation effect, fit the model, test the additional effect for significance, and repeat for all possible parameters to add to the model. We performed this procedure for both the small friendship and the senate collaboration data. The results for significant effects are shown in Figure~\ref{fig:sigtesting}. We see in both the friendship data and the senate data results that most of the significant effects have absolute value less than ten. In addition, the $p$-values for the effects from the friendship data are more spread out than the $p$-values for the senate data, which are concentrated at about 0.02 or less. This may suggest that larger data sets tend to result generally in smaller $p$-values, just like a larger sample size results in smaller $p$-values in a $t$-test.}

@@

We have varied model settings already by choosing models M1, M2, M3 to fit to our small friendship network data set. In Section~\ref{sec:samemodsamedat}, we fit these models to the data 1,000 times, and in this section, we explore simulations from these three models given the mean values of from the 1,000 fitted parameter values as the parameters in our models. From each of these three models using the means of parameter estimates as our fixed parameter values, we simulated 1,000 observations from each of the three models. In this process, we condition on the first friendship network observation, and the second and third observations are simulated from the SAOM models with the given parameter values. From these simulations, we first create a visualization that represents an average network.

@@

To create the average network visualization shown in Figure~\ref{fig:averagenet}, we follow the same procedure as in Section~\ref{sec:minds}, counting occurrences of each possible edge in the simulations, resulting in a summary network with weighted edges representing the number of times an edge appeared in the simulated wave 2 when simulating from the SAOM 1,000 times. As in Figure~\ref{fig:summnetM1}, edges only appear in the average networks if they appear more than 5\% of the time in the simulations. In Figure~\ref{fig:averagenet}, we show the ``average'' network from the three models we fit and the first and second waves of data. \st{Comparing the three averages to waves 1 and 2, we see that they have very similar structure to wave 1. Model 2, which included the transitive triplet parameter, seems to have created a larger connected component overall than models 1 and 3. In particular, if we look at the group of nodes $\{10,11,14\}$, we see they are very strongly connected within the three average networks, and they are completely separate from the other nodes in the true wave 2. None of the three average networks show node 16 gaining ties as it does in wave two, nor do they show nodes 4 and 7 becoming isolated. In Model 2, however, the ties to node 7 appear much weaker than in Model 1 or Model 2, suggesting that of the three, Model 2 may be the best fit for our data.}  %XXXX ST commented out after I realized I had made the same mistake as figure 7. XX This could be an indication that, despite the highly significant nature of the parameter $\beta_3$ according to the $t$-tests in \texttt{RSiena}, the transitive triplet parameter should not be included because it does not result in a structure similar to the data.

@@

\subsubsection{Fitting the same model to different data}\label{sec:samemoddiffdat}

\st{As mentioned in Section~\ref{sec:models}, we fit the models M1, M2, and M3 to both the small friendship network and the senate collaboration network. We fit each model to the friendship data 1,000 times, and to the senate data 100 times. The means and standard deviations of the parameter estimates for each combination of model and data are given in Table~\ref{tab:meanests} }

% table of fitted M1 M2 M3
\begin{table}
\scalebox{0.9}{
\begin{tabular}{l|ccc|ccc }
  \hline
  & \multicolumn{3}{|c|}{Friendship Data} & \multicolumn{3}{|c|}{Senate Data} \\
  \hline
  & M1 & M2 & M3 & M1 & M2 & M3 \\
  $\alpha_1$ & \Sexpr{SFE$M1[1]} & \Sexpr{SFE$M2[1]} & \Sexpr{SFE$M3[1]} & \Sexpr{SSE$M1[1]} & \Sexpr{SSE$M2[1]} & \Sexpr{SSE$M3[1]} \\
  $\alpha_2$ & \Sexpr{SFE$M1[2]} & \Sexpr{SFE$M2[2]} & \Sexpr{SFE$M3[2]} & \Sexpr{SSE$M1[2]} & \Sexpr{SSE$M2[2]} & \Sexpr{SSE$M3[2]}  \\
  $\alpha_3$ & -- & -- & -- & \Sexpr{SSE$M1[3]} & \Sexpr{SSE$M2[3]} & \Sexpr{SSE$M3[3]} \\
  $\beta_1$ & \Sexpr{SFE$M1[3]} & \Sexpr{SFE$M2[3]} & \Sexpr{SFE$M3[3]} & \Sexpr{SSE$M1[4]} & \Sexpr{SSE$M2[4]} & \Sexpr{SSE$M3[4]} \\
  $\beta_2$ & \Sexpr{SFE$M1[4]} & \Sexpr{SFE$M2[4]} & \Sexpr{SFE$M3[4]} & \Sexpr{SSE$M1[5]} & \Sexpr{SSE$M2[5]} & \Sexpr{SSE$M3[5]}\\
$\beta_3$ & -- & \Sexpr{SFE$M2[5]} & -- & -- & \Sexpr{SSE$M2[6]} & -- \\
  $\beta_4$ & -- & --  & \Sexpr{SFE$M3[6]} & -- & -- & \Sexpr{SSE$M3[7]} \\
  \hline
\end{tabular}
}
\caption{\label{tab:meanests} The means (standard deviations) of parameter values estimated from repeated fittings of $M1, M2, M3$ to the small friendship network and the senate collaboration network. Each model was fit 1,000 times to the friend data, while each model was fit 100 times to the senate data.}
\end{table}

<<compareModels, fig.cap="Density plots of the repeated estimates from fitting models M1, M2, and M3 to the friendship and senate example data. Note that the rate parameters are excluded since the two data sets have different numbers of waves. The first two parameters, outdegree and reciprocity, have the same relationship for both data sets: the outdegree parameter estimate is strongly negative, while the reciprocity estimate is strongly postitive. For the friendship data, the inclusion of the transitive triplet parameter has strong effect on the estimates of the other two parameters, while it does not affect the estimates of the other two parameters for the senate data.", fig.height = 3>>=
simu2$data <- "friends"
senateEsts$data <- "senate"
allDataEsts <- rbind(senateEsts, simu2)
allDataEsts$parameter <- as.factor(allDataEsts$parameter)
allDataEsts$parameter <- ordered(allDataEsts$parameter, levels = c(paste0("alpha", 1:3),paste0("beta", 1:4)))
ggplot(allDataEsts %>% filter(str_detect(parameter, "beta"))) + 
  geom_density(aes(x = estimate, fill = Model), alpha = .5) + 
  facet_grid(data~parameter, scales = "free", labeller = "label_both") + 
  ThemeNoNet + 
  theme(legend.position = 'bottom')
@

\st{Looking at Figure~\ref{fig:compareModels} and Table~\ref{tab:meanests}, we see a few patterns in the estimates from both models. First, we can see in both the table and the density of the estimates that the same relationship between the outdegree parameter, $\beta_1$, and the reciprocity parameter, $\beta_2$. In both data sets and across all three models, the estimates of $\beta_1$ are all negative and hover between -5 and -3, while the estimates of $\beta_2$ are all positive and hover between four and five. This suggests that in both data sets, nodes are \textit{discouraged} from forming ties that are outgoing without being reciprocated, while also being \textit{encouraged} to form outgoing ties to nodes that tie to them. In both data sets, people seem to want to have reciprocated relationships: teenage girls want be friends with other girls that reciprocate their friendship, and senators want to coauthor bills with senators who have also been coauthors on their bills. We explore the relationship between $\beta_1$ and $\beta_2$ further in Section~\ref{sec:samemodsamedat}.

The inclusion of $\beta_3$, the jumping transitive triplet parameter, for the friendship data had a noticeable effect on the other parameters in the model. The same cannot be said for the inclusion of $\beta_3$ for the senate data. The covariate used in the senate data for the jumping transitive triplet calculation was the number of bills authored by the ego node in the given year. The number of bills authored by a senator varies wildly, from no bills to 114 bills authored in two years, so this effect could simply be nonsensical for the senate data, since senators are less likely to have the same number of bills authors than teenage girls are to have the same drinking behavior. Looking at the estimates of $\beta_4$, we see that the estimates for the senate data are near zero, suggesting this effect, which considers indirect ties, is not important for the senate data: indirect relationships do not describe the senate collaboration structure as much as they do the teenage friendship structure. }

@@

\subsubsection{Fitting the same model to the same data}\label{sec:samemodsamedat}

<<distests, fig.cap="Distribution of fitted parameter values for our three SAOMs. The inclusion of $\\beta_3$ or $\\beta_4$ clearly has an effect on the distribution of the rate parameters, $\\alpha_1$ and $\\alpha_2$.", fig.height=3>>=

simu2 <- read.csv("data/simulation-1000-M1-M2-M3.csv")

@@

<<getdatam, eval = FALSE>>=
nulls <- read.csv("data/distribution_null_model.csv")
alt <- read.csv("data/distribution_jumpTT_model.csv")
alt2nd <- read.csv("data/distribution_dblpairs_model.csv")
means <- simu2 %>% group_by(Model, parameter) %>% summarize(
  means = mean(estimate)
)
null_mod_sv <- as.numeric(nulls[,-c(1,6,7)] %>% summarise_each(funs(mean)))
alt_mod_sv <- as.numeric(alt[,-c(1, 7,8)] %>% summarise_each(funs(mean)))
alt_mod2_sv <- as.numeric(alt2nd[,-c(1,7,8)] %>% summarise_each(funs(mean)))
@

@@

To further investigate the odd relationship between the parameter values, we look at correlations between each of the parameter estimates in each model. In Figure~\ref{fig:corplots}, we examine correlations between each of pair of parameters within each model and overall.\nocite{ggally} The strongest correlation within each model is between $\beta_1$ and $\beta_2$, with absolute value of correlation between those two parameter values greater than 0.90 in all three models. The $\beta_1$ parameter is also highly correlated with the $\beta_3$ parameter within model M2, but it is not as highly correlated with the $\beta_4$ parameter in model M3. \st{XX NEED TRANSITION XX }

@@

\st{The last principle of model visualization is to explore the process of fitting the model, instead of just focusing on the end result. This principle is perhaps the most important for SAOMs because} the model fitting process in \texttt{RSiena} involves several simulation steps that are hidden from the user. Hiding the MCMC steps is practical and efficient if a researcher is primarily interested in fitting one model to a set of longitudinal network data, obtaining parameter estimates, \st{and drawing conclusions or making predictions}. We are more interested in \textit{how} the models are fit, so we extracted and explored the different steps of that process instead of allowing them stay hidden.

A key component of each step of the SIENA method of moments algorithm is the ``microstep" process. A series of microsteps is obtained by simulating from the model in its current state, $x(t_m)$ with current parameter values $\mathbf{\theta}_{0} = \{\alpha_{1_0}, \dots \alpha_{{m-1}_0}, \beta_{1_0}, \dots, \beta_{K_0}\}$, to the next state, $x(t_{m+1})$. This microstep process stops when the simulated network has achieved the same number of differences, $C$, from $x(t_m)$ as $x(t_{m+1})$, where 
$$C = \sum_{i \neq j} |x_{ij}(t_{m+1}) - x_{ij}(t_{m})|.$$ 
This simulation process follows the steps of the continuous-time Markov chain. \st{Each tie change in the CTMC is referred to as one ``microstep". At each microstep}, an ``ego node" is selected to make a change, and the chosen ego node randomly makes one change in its ties according to the probabilities, $\{p_{ij}: i\neq j \in \{1, \dots, n\}\}$ defined in Equation~\ref{eq:pij}, determined by its objective function. The options for change are (1) removing a current tie, (2) adding a new tie, or (3) making no change at all. Saving and exploring all of these steps is not computationally efficient if one is only interested in estimating parameter values, but they can be saved and extracted using options in \texttt{RSiena}, which is what we did to create our visualizations. Between two network observations $x(t_m)$ and $x(t_{m+1})$, there can be dozens, hundreds, or even thousands of microsteps, depending on the size of the network and the number of changes between two network observations. We wanted to view these in-between steps in order to better understand the behavior of the underlying continuous-time Markov chain.

The first vizualization we present here is an animation of the simulated microsteps that form the transition steps of the CTMC from wave 1 to wave 2 of the small friendship network example shown in Figure~\ref{fig:alldataviz} when fitting model M1. Movies similar to this animation were used to visualize the changes of dynamic networks in \citet{dynnetviz}. When each ego node is selected in a microstep, it is emphasized in the animation, then the associated edge either appears or disappears. If there are no changes at a particular microstep, no changes are seen. Some frames of the animations are shown in Figure~\ref{fig:makinggif}, and the full movie can be viewed at INSERT LINK TO GIF.

@@

set.seed(34569234)
froms <- static_pall %>% select(id = from, microstep, x = from.x, 
        y = from.y, ms, .frame, size, color) %>% unique
tos <- static_pall %>% select(id = to, microstep, x = to.x, y = to.y, 
        ms, .frame) %>% unique
tos <- left_join(tos, froms)
nodedata <- rbind(froms, tos) %>% unique
nodedata <- nodedata %>% replace_na(replace = list(size = 1, 
        color = "#333333"))
to_ani <- ggplot() + 
  geom_curve(data = static_pall, aes(x = from.x, 
        y = from.y, xend = to.x, yend = to.y, frame = .frame, 
        colour = ecolor, size = esize), curvature = 0.1, arrow = arrow(angle = 20, 
        length = unit(0.1, "inches"))) + 
  geom_point(data = nodedata, 
        aes(x = x, y = y, colour = color, size = size, frame = .frame)) +
  scale_colour_identity() + scale_size_identity() + theme_void() + 
        coord_fixed()
to_ani + 
  facet_wrap(~.frame, nrow=2, labeller = 'label_both') + 
  ThemeNet + 
  theme(panel.background = element_rect(color = 'black'))
@
